---
title: "Introduction to saemMicrobiome"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to saemMicrobiome}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(saemMicrobiome)
```

## What problem this package solves

Longitudinal microbiome data (several samples per subject over time) usually
have two characteristics that standard mixed models do not handle well:

1. **Zero inflation**: many taxa are absent in a large fraction of the
   samples, in addition to having a continuous distribution conditional on
   being present.
2. **Dependence between observations of the same subject**: a per-subject
   random effect must be modelled, not only fixed effects.

`saemMicrobiome` offers two models for this problem, both estimated with the
Stochastic Approximation EM (SAEM) algorithm developed by John Barrera:

- **ZIBR** (`fit_zibr()`): for when the response is a **proportion** (for
  example, the relative abundance of a taxon with respect to the sample
  total), in the interval `[0, 1)`.
- **ZIBBMR** (`fit_zibbmr()`): for when the response is a **count** (number of
  reads of the taxon) together with the sample's total reads (sequencing
  depth). It is preferable to ZIBR when the sequencing depth varies a lot
  across samples, because it models that variation explicitly instead of
  working only with the already-normalized proportion.

## Simulate data and fit a ZIBR model

We use 300 subjects, a sample size large enough for the estimates to land
close to the true values of the simulation (`alpha = c(-0.3, 0.5)`,
`beta = c(0.2, -0.4)`, `phi = 15`).

```{r zibr-basico}
# You only need to change n_subjects / n_time; n_obs is derived from them.
n_subjects <- 300
n_time <- 4
n_obs <- n_subjects * n_time

set.seed(3)
dat <- simulate_zibr_data(
  n_subjects = n_subjects, n_time = n_time,
  X = matrix(rbinom(n_obs, 1, 0.5)), Z = matrix(rbinom(n_obs, 1, 0.5)),
  alpha = c(-0.3, 0.5), beta = c(0.2, -0.4),
  sigma_alpha = 0.4, sigma_beta = 0.3, phi = 15, seed = 3
)

head(dat)

fit <- fit_zibr(
  y = dat$Y, id = dat$Subject, X = dat$X.1, Z = dat$Z.1,
  phi_start = 10, alpha_start = c(-0.2, 0.1), beta_start = c(0.1, 0.1),
  n_iter = 500, seed = 1, compute_fim = TRUE
)

fit
```

The returned object has standard R methods for fitted models:

```{r zibr-metodos}
coef(fit)
logLik(fit)
se(fit)
```

> **Note:** with some data sets, or with few iterations, the information matrix
> of the logistic part can be ill-conditioned and `se()` can return `NaN` for
> those coefficients. It does not affect the estimates (the coefficients are
> correct); it only indicates that the standard error of those terms could not
> be computed. It is usually resolved by increasing `n_iter` or the number of
> subjects.

The `plot()` method offers several plots depending on the `which` argument:
convergence of the algorithm, coefficients with confidence intervals, and the
distribution of the random effects.

```{r zibr-plot-conv, fig.width=7, fig.height=5}
plot(fit, which = "convergence")    # trace of the algorithm (default)
```

```{r zibr-plot-coef, fig.width=7, fig.height=4}
plot(fit, which = "coefficients")   # coefficients with 95% CI
```

```{r zibr-plot-ranef, fig.width=7, fig.height=3.5}
plot(fit, which = "random")         # random effects per subject
```

There are also fit-diagnostic plots: observed vs. predicted and residuals (of
the continuous part, on the observations where the taxon is present).

```{r zibr-plot-fit, fig.width=6, fig.height=5}
plot(fit, which = "fit")            # observed vs. predicted
```

```{r zibr-plot-resid, fig.width=7, fig.height=3.5}
plot(fit, which = "residuals")      # residuals
```

## The same workflow for ZIBBMR (counts)

When you have the read count and the total reads per sample, `fit_zibbmr()`
models that information directly with a beta-binomial likelihood, instead of
working with the already-divided proportion:

```{r zibbmr-basico}
# Reuses n_subjects / n_time / n_obs from the previous example.
S <- rep(1000, n_obs)
set.seed(3)
dat_counts <- simulate_zibbmr_data(
  n_subjects = n_subjects, n_time = n_time, S = S,
  X = matrix(rbinom(n_obs, 1, 0.5)), Z = matrix(rbinom(n_obs, 1, 0.5)),
  alpha = c(-0.3, 0.5), beta = c(0.2, -0.4),
  sigma_alpha = 0.4, sigma_beta = 0.3, phi = 15, seed = 3
)

fit_counts <- fit_zibbmr(
  y = dat_counts$Y, S = dat_counts$TotalCounts, id = dat_counts$Subject,
  X = dat_counts$X.1, Z = dat_counts$Z.1,
  phi_start = 10, alpha_start = c(-0.2, 0.1), beta_start = c(0.1, 0.1),
  n_iter = 500, seed = 1, compute_fim = TRUE
)

fit_counts
```

## Fit several taxa of a data frame

In practice the data come as a wide data frame (one column per taxon).
`fit_zibr_taxon()`/`fit_zibbmr_taxon()` avoid having to build the `X`/`Z`
matrices for each taxon by hand, and `fit_zibr_taxa()`/`fit_zibbmr_taxa()`
let you fit several taxa at once:

```{r taxones}
sim <- simulate_microbiome_data(n_ind = 15, n_time = 4, n_taxa = 3, seed = 1)
head(sim$proportion)

fits <- fit_zibr_taxa(
  data = sim$proportion, taxa = sim$taxa, id = "id",
  covariates = c("time", "group"), n_iter = 50, seed = 1
)

sapply(fits, coef)
```

## Compare nested models with a likelihood-ratio test

To decide whether a covariate contributes significantly, compare a full model
against a reduced one (without that covariate) with `lrt_zibr()`/
`lrt_zibbmr()`:

```{r lrt}
full <- fit_zibr_taxon(
  data = sim$proportion, taxon = "Taxon1", id = "id",
  covariates = c("time", "group"), n_iter = 50, seed = 1
)
reduced <- fit_zibr_taxon(
  data = sim$proportion, taxon = "Taxon1", id = "id",
  covariates = "time", n_iter = 50, seed = 1
)

lrt_zibr(full, reduced, df = 1)
```

For many taxa at once, `lrt_zibr_table()`/`lrt_zibbmr_table()` build a table
with one row per taxon.

## The "mother" function: fit_saem_microbiome()

If you prefer not to choose between `fit_zibr()`/`fit_zibbmr()` explicitly,
`fit_saem_microbiome()` (an alias of `fit_microbiome_model()`) dispatches to
the right model according to the `model` argument, taking the response, the
total reads (if applicable) and the covariates directly from a data frame by
column name:

```{r madre}
fit_saem_microbiome(
  model = "zibr", data = sim$proportion, taxon = "Taxon1",
  id = "id", covariates = c("time", "group"), iter = 50, seed = 1
)
```

## Your own data: preparing a Romero-type data set

If the data come in the format of the Romero study (a list with `SampleData`
and `OTU`), `prepare_romero_zibr()`/`prepare_romero_zibbmr()` build the
standard covariates (scaled gestational time, scaled age, time-pregnancy
interaction) and filter taxa by zero proportion, leaving everything ready for
`fit_zibr_taxon()`/`fit_zibbmr_taxon()`. See `?prepare_romero_zibr` for the
expected format of `SampleData`/`OTU`.
